\vspace{-2.0cm}The effect of nitrogen application time on infection and development of the Wheat Streak Mosaic Virus (WSMV)

### 2015-09-29 draft1leslieandkenny.Rnw
Added some simulated data and a bit about the model.

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}

@@

@@

@@

\section{Statistical Experience and Assistance Needed}
\noindent Nar B. Ranabhat is a PhD candidate in the Department of Land Resources and Environmental Science. He is currently working on several projects regarding the spread and development of the Wheat Streak Mosaic Virus (WSMV) in Winter Wheat. Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently enrolled in Mixed Models. Nar has asked for our help in building a model, fitting it in R, and interpreting the results.

@@

Nar conducted this experiment in a MSU field at the base of the Bridger Mountains. He chose one area in the field and divided it into six $31.5$ by $27.5$ meter blocks. The blocks were then divided into five rows, with $5$ meters of space between each row. Each row was randomly assigned to a variety of Winter Wheat with separate randomizations in each block. The rows then were divided into six $1.5$ by $5$ meter plots. Each plot was randomly assigned to one of six combinations of nitrogen application time and inoculation status, with separate randomizations in each row. The combinations were fall/inoculated, fall/control, early spring/inoculated, early spring/control, late spring/inoculated, and late spring/control. All plots were planted in the {\bf FALL?} and allowed to grow over the winter. The following {\bf September?}, $30$ plants were chosen from each plot to be taken to the lab and tested for presence of the virus. The selection of the plants from each plot was not random nor technically systematic. The {\bf undergraduate?} technicians were instructed to collect a sample of plants spread evenly throughout the field. {\bf different undergraduates doing picking? picking done all at the same time? is there visual evidence of the disease?} A total of $180$ plants were collected. The plants were then sent to the lab and screened for the virus via the ELISA procedure. This is a three year study. The data are collected for the first year, and the second year plants are waiting to be analyzed in the lab {\bf describe how they are stored during waiting time}. Planting for the third year will begin this fall {\bf I think?}.  \\

@@

@@

<<interactionplot,echo=FALSE>>=
qplot(x = n.trt, y = infected, geom = 'boxplot', color = status,
      facets = .~variety, data = fert2) +
  theme(axis.text.x = element_text(angle = 90))
@

@@

# Order the dataset
#fert.ordered <- arrange(fert2, block, variety, status, n.trt)
fert.ordered <- arrange(fert2, variety, block, status, n.trt)

@@

# Plot the heatmap, with zeros in black and segments separating the blocks
par(mar = c(5, 10, 6, 2))
image(x = seq(0.5, 5.5, 1/6), y = 1:6, z = infected.arranged,
      col = c('black', heat.colors(30)), xlab = '', ylab = '',
      zlim = c(0, 30), yaxt = 'n', xaxt = 'n')
title('Infection Counts', line = 4)
segments(x0 = 1.5:4.5, y0 = 0.5, y1 = 6.5, col = 'white')
axis(3, labels = rep(levels(fert2$block), 5),
     at = seq(7/12, 5 + 5/12, 1/6), cex.axis = 0.75)
axis(2, labels = levels(fert2$n.status), at = 1:6, las = 2)
axis(1, labels = levels(fert2$variety), at = 1:5)

@@

<<models,cache=TRUE>>=
glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status +
                    (1|block/row),
                  control = glmerControl(optimizer = 'bobyqa'),
                  family = binomial, data = fert2)
summary(glmm3way)

glmm2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status +
                    n.trt*status + (1|block/row),
                    control = glmerControl(optimizer = 'bobyqa'),
                  family = binomial, data = fert2)
summary(glmm2way)

@@

<<plots1,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
par(mfrow = c(1, 2))
plot(resid(glmm3way) ~ predict(glmm3way), col = variety, data = fert2)
plot(resid(glmm3way) ~ fitted(glmm3way), col = variety, data = fert2)
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert2$variety))
@

>>>>

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}

@@

# If status=INOC is the most interesting case, make INOC the baseline
fert2$status2 <- relevel(fert2$status, 'INOC')

@@

<<fakedata,echo=FALSE>>=
# Create some fake data with no differences anywhere
set.seed(983724)
fert.sim1 <- data.frame(
  'block' = c(rep('A', 30), rep('B', 30), rep('C', 30), rep('D', 30), rep('E', 30), rep('F', 30)),
  'variety' = c(rep('Var1', 6), rep('Var2', 6), rep('Var3', 6), rep('Var4', 6), rep('Var5', 6)),
  'n.application' = rep(c('FALL', 'FALL', 'EARLY SPRING', 'EARLY SPRING', 'LATE SPRING', 'LATE SPRING'), 30),
  'status' = rep(c('CNTL', 'INOC'), 90),
  'total' = rep(30, 180),
  'infected' = sample(fert2$infected, 180, replace = TRUE))

# Correctly order nitrogen application levels
fert.sim1$n.trt <- relevel(fert.sim1$n.application, 'FALL')

# If status=INOC is the most interesting case, make INOC the baseline
fert.sim1$status2 <- relevel(fert.sim1$status, 'INOC')

# Create a row factor
fert.sim1$row <- with(fert.sim1, interaction(variety, block))
@

@@

\section{Statistical Experience and Assistance Needed}
\noindent Nar B. Ranabhat is a PhD candidate in the Department of Land Resources and Environmental Science. He is currently working on several projects regarding the spread and development of the Wheat Streak Mosaic Virus (WSMV) in Winter Wheat. Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently auditing Statistics $448$. Nar has asked for our help in building a model, fitting it in R, and interpreting the results.

@@

Nar conducted this experiment in a MSU field at the base of the Bridger Mountains. He chose one area in the field and divided it into six $31.5$ by $27.5$ meter blocks. The blocks were then divided into five rows, with $5$ meters of space between each row. Each row was randomly assigned to a variety of Winter Wheat with separate randomizations in each block. The rows then were divided into six $1.5$ by $5$ meter plots. Each plot was randomly assigned to one of six combinations of nitrogen application time and inoculation status, with separate randomizations in each row. The combinations were fall/inoculated, fall/control, early spring/inoculated, early spring/control, late spring/inoculated, and late spring/control. All plots were planted in the {\bf FALL?} and allowed to grow over the winter. The following {\bf September?}, $30$ plants were chosen from each plot to be taken to the lab and tested for presence of the virus. The selection of the plants from each plot was not random nor technically systematic. The {\bf undergraduate?} technicians were instructed to collect a sample of plants spread evenly throughout the field. {\bf different undergraduates doing picking? picking done all at the same time? is there visual evidence of the disease?} A total of $180$ plants were collected. The plants were then sent to the lab and screened for the virus via the ELISA procedure. This is a three year study. The data are collected for the first year, and the second year plants are waiting to be analyzed in the lab {\bf describe how they are stored during waiting time}. Planting for the third year will begin this fall {\bf I think?}.\\

@@

Because the response variable is a binomial count, and because the design
includes random assignment to rows within blocks, a binomial generalized
linear mixed model is appropriate. This model will estimate the probability
of infection for each \verb|variety:treatment:status| combination while
controlling for differences between blocks and between
rows in the same block.

The model is
$$
y_i\sim\mathrm{Binomial}(30, p_i),
$$
$$
\mathrm{logit}(p_i)=\mathbf{x}_i\mathbf{\beta}+b_{0,j[i]}+b_{1,j[i],k[i]}
$$
where
\begin{itemize}
\item \(p_i\) is the probability of infection for the \(i\)th observation,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection,
\item \(\mathbf{x_i}\) is a row vector containing indicator variables for
the variety, nitrogen application timing, and inoculation status of
observation \(i\), and
\item \(b_{0,j}\sim\mathrm{N}(0,\sigma^2_{b_0})\) and
\(b_{1,j,k}\sim\mathrm{N}(0,\sigma^2_{b_1})\) are random effects for block
and row, respectively.
\end{itemize}

@@

<<interactionplot,echo=FALSE>>=
qplot(x = n.trt, y = infected, geom = 'boxplot', color = status,
      facets = .~variety, data = fert.sim1) +
  theme(axis.text.x = element_text(angle = 90))
@

@@

# Order the dataset
fert.ordered <- arrange(fert.sim1, variety, block, status, n.trt)

@@

# Plot the heatmap, with zeros in black and segments separating the blocks
par(mar = c(5, 10, 6, 2))
image(x = seq(0.5, 5.5, 1/6), y = 1:6, z = infected.arranged,
      col = c('black', heat.colors(30)), xlab = '', ylab = '',
      zlim = c(0, 30), yaxt = 'n', xaxt = 'n')
title('Infection Counts', line = 4)
segments(x0 = 1.5:4.5, y0 = 0.5, y1 = 6.5, col = 'white')
axis(3, labels = rep(levels(fert.ordered$block), 5),
     at = seq(7/12, 5 + 5/12, 1/6), cex.axis = 0.75)
axis(2, labels = levels(fert.ordered$n.status), at = 1:6, las = 2)
axis(1, labels = levels(fert.ordered$variety), at = 1:5)

@@

<<models,cache=TRUE>>=
glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status +
                    (1|block/row),
                  control = glmerControl(optimizer = 'bobyqa'),
                  family = binomial, data = fert.sim1)
summary(glmm3way)

glmm2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status +
                    n.trt*status + (1|block/row),
                    control = glmerControl(optimizer = 'bobyqa'),
                  family = binomial, data = fert.sim1)
summary(glmm2way)

@@

<<plots1,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
par(mfrow = c(1, 2))
plot(resid(glmm3way) ~ predict(glmm3way), col = variety, data = fert.sim1)
plot(resid(glmm3way) ~ fitted(glmm3way), col = variety, data = fert.sim1)
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert.sim1$variety))
@
